\section{Virgil's {\AE}neid:  Lines 1-33} % (fold)



\latline
  {[1] \={\macron A}rm\-a v\-ir\=umqu\-e c\-an{\={\macron o}}, $||$ Tr\={o}j\={\ae} qu\={\macron{\i}} pr\={\macron{\i}}m\-us \-ab \={o}r\={\macron{\i}}s}
	{I sing of arms and the man, who came from Troy to}
  {180}
	{I sing of arms and the man, an exile, who came from the borders of Troy,}
	{
	\begin{enumerate}
			\item The first word is important, the use of ``arma'' indicates that Vergil is singing about war.
			\item Vergil's first word and last words of sentences and clauses are carefully chosen.
			\item Supply ``est'':  ``{\AE}neas primus est; profugus est''
			\item There are \textbf{three} types of adjectives
			\begin{enumerate}
				\item attributive: simple modification ``vir bonus''
				\item substantive:  ``bonus'' i.e. ``the good man''
				\item predicate:  ``vir bonus est''
			\end{enumerate}
		 \end{enumerate}	
	}

\latline
		{{\macron I}t\-{a}l\-i\={a}m f\={\macron a}t\={\macron o} pr\-of\-ug\={\macron u}s $||$ L\={\macron a}v\={\macron{\i}}n\-aqu\-e v\={\macron e}n\={\macron{\i}}t}
		{Italy and the Lavinian shores under an exile's fate.}
		{80}{to Italy and the Lavinian }{}

\newpage
\latline
{l\={\macron{\i}}t\-or\-a --- m\=ultu\sout{m }\={i}ll\={{\sout{e }}\=e}t $||$ t\=err\={\macron{\i}}s j\=act\={\macron a}t\-us \-et \=alt\={\macron o}}
{--- after he was buffeted both by many lands and also}
{110}
{shores, --- that man, much buffeted both on land and sea, as well as by}
{}


\latline
{v\={\i} s\-up\-er\=um, s\={\ae}v\={\ae} $||$ m\-em\-or\=em J\={u}n\={o}n\-is \-ob \={\i}r\=am,}
{the powers of the Gods, brought to rage by the cruel memory of Juno}
{110}
{the powers of the gods, on account of the unforgetting memory of wrathful Juno,}
{
	\begin{enumerate}
		\item note the meter:  ``we-superroom''
		\item Note the interlocked word order:  as written: ``saevae memorem J\={u}n\={o}nis ob {\={\i}}ram'',
		but has the force of  ``saevae {\={\i}}ram J\={u}n\={o}nis ob \={\i}ram memorem''
	\end{enumerate}
}

\latline
{m\=ult\-a qu\-oqu\={\sout{\=e\= }\=e}t b\=ell\={o} $||$ p\=ass\=us, d\=um c\=ond\-er\-et \=urb\=em}
{endured many things and war, thence would found the city}
{110}
{he also endured many things in war, that he might found the city}
{
\begin{enumerate}
	\item ablative of respect ``bell{\macron o}''
	\item due to the subjunctive, we say ``might.''  This is a \textbf{dum} clause
	\item While we made ``jactatus est'' passive, ``passus'' remains active in sense as it is a \textbf{deponent} verb
\end{enumerate}
}

\newpage

\latline
{\={\macron{\i}}nf\=err\=etqu\-e d\-e\={o}s L\-at\-i\={o} $||$ --- g\-en\-us \=und\-e L\-at\={\i}n\=um}
{and carry the gods to Latinum --- and sire the Latin race,}
{110}
{and might carry the gods to Latium --- from which place (whence) [are] the}
{
\begin{enumerate}
	\item Lati{\macron o} is a dative of direction, this is commonly used with compound verbs
	\item ``unde'' is a relative pronoun
\end{enumerate}
}


\latline
{\=Alb\={a}n\={\macron {\i}}qu\-e p\-atr\={e}s $||$ \-atqu\sout{e }\=alt\={\ae} m\={oe}n\-i\-a R\={\macron o}m\={\ae}.}
{the Alban forefathers and also the high walls of Rome.}
{110}
{Alban fathers and also the walls of mighty Rome.}
{
\begin{enumerate}
	\item Good demonstration of long or short by position: ``patr{\macron
e}s''. You have `a' followed by `tr', thus seeming to indicate long by
position. But since `r' may or may not count, this is not definite. It, in
fact, remains short.
\end{enumerate}
}

\latline
{M\={\macron u}s\-a, m\-ih\={\i} c\={au}s\={a}s m\-em\-or\={a}, $||$ qu\={o} n\={u}m\-in\-e l\={\ae}s\={o}}
{Muse, recount to me the causes by which, divinity having been offended or  }
{110}
{Muse, recall to me the causes by which }
{
\begin{enumerate}
	\item n\={u}mine l{\ae}s\=o: ablative absolute
\end{enumerate}
}

\newpage


\latline
{quidve dol\={\macron e}ns r\={\macron e}g\={\macron{\i}}na deum tot volvere c\={\macron a}s\={\macron u}s}
{or the vexation at what would impelled the queen of the gods to force}
{110}
{or vexation at what impelled the queen of the gods to spin out such harsh fates and }
{
\begin{enumerate}
	\item Indirect question set up by ``quid''
	\item IQ necessitates subjunctive, thus ``impulerit''
\end{enumerate}
}


\latline
{\={\macron{\i}}ns\=ign\=em p\-i\-et\={\macron a}t\-e v\-ir\=um, $||$ t\-ot \-ad\={\macron{\i}}r\-e l\-ab\={\macron o}r\={\macron e}s }
{so many difficult labors}	
{110}
{and bring that man, outstanding in piety, to face so many harsh labors.}
{}

\latline
{\=imp\-ul\-er\=it. $||$ T\=ant\={\ae}n\sout{e }\=anim\={\i}s c\={\ae}l\=est\-ib\-us \={\macron{\i}}r{\ae}?}
{Are the minds of the heavens so wrathful?}
{110}
{}
{
\begin{enumerate}
	\item Note ``tantae'' is a predicate adjective
\end{enumerate}
}

\newpage

\latline
{\=Urbs \=ant\={\macron{\i}}qu\-a f\-u\=it (T\=yri\={\macron{\i}} t\-en\-u\={\macron e}r\-e c\-ol\={\macron o}n\={\macron{\i}})}
{It was an ancient city (founded Tyrian colonists)}
{100}
{}
{}



\latline
{K\=arth\={\macron a}g\={\macron o}, \=Ita\-l\-i\=am c\=ontr\={\macron a} T\-ib\-er\={\macron{\i}}n\-aqu\-e l\=ong\={\macron e}}
{Carthage, facing Italy and the entryway of the Tiber at a }
{110}
{}
{}

\latline
{\={\macron o}st\-i\-a, d\={\macron{\i}}v\-es \-op\=um st\-ud\-i\={\macron{\i}}squ\-e \-asp\=err\-im\-a b\=ell\={\macron{\i}};}
{distance, rich in wealth and zeal for harshest war}
{110}
{}
{}


\newpage
\latline
{qu\=am J\={\macron u}n\={\macron o} f\=ert\=ur t\=err\={\macron{\i}}s m\-ag\-is \=omn\-ib\-us \={\macron u}n\=am}
{a city which Juno, it is said, loved above all lands, only}
{110}
{}
{}



\latline
{p\=osth\-ab\-it\={\macron a} c\-ol\-u\=iss\-e S\-am\={\macron o}: h\={\macron{\i}}c \=ill\-i\-us \=arm\=a,}
{esteeming Samos less:  here were her weapons and}
{110}
{}
{\begin{enumerate}
	\item ``h\={\i}c'' indicates ``here'', ``hic'' indicates ``this man''
\end{enumerate}}

\latline
{h\={\macron{\i}}c c\=urr\=us f\-u\-it; h\=oc r\={\macron e}gn\=um d\-e\-a g\=ent\-ib\-us \=ess\=e,}
{her chariot, here the goddess already then aims and favors that this be a kingdom for clans}
{160}
{her chariot, that this city would be the ruling power,}
{\begin{enumerate}
	\item h\=oc r\=egn\=um is the subject accusative
	\item esse is the infinitive
	\item there are three types of infinitive:
		\begin{enumerate}
			\item complementary infinitive ``debe\=o ven\=ire''
			\item indirect statement
			\item nominative gerund ``err\=are humanum est''
		\end{enumerate}
\end{enumerate}}

\newpage

\latline
{s\={\macron{\i}} qu\={\macron a} f\={\macron a}t\-a s\-in\=ant, j\=am t\=um t\=end\=itqu\-e f\-ov\=etqu\=e.}
{where, if the Fates were willing, she would bless and favor.}
{290}
{if the Fates would permit it, further, she long held and nourished.}
{\begin{enumerate}
	\item ``after si, nisi, numine, `ali' takes a holiday''
	\item Conditions Review
	\begin{enumerate}
		\item Simple / General Conditions
		\begin{enumerate}
			\item Use Indicative Mood
			\item Present == Pr. Indicative + Pr. Indicative: s\=i Aeneas vincit, beatus est:  If A. wins, he is happy
			\item Perfect / Pluperfect == Perfect Indicative + Perfect Indicative:  s\=i Aeneas vinxit, fuit beatus:  If A won, he was happy
			\item Future ( aka \textbf{Future More Vivid}) == Futur Indict + Future Indic:  S\=i A vincet erit beatus:  If A wins, he will be happy
		\end{enumerate}
		\item Contrafactual 
		\begin{enumerate}
			\item Present Contrafactual == Imp. Subj: S\=i A. vincerent, esset beatus: If A were to win, he would be happy
			\item Past Contrafactual == Pluperf Subj:  S\=i A. vinxisset, fuisset beatus:  If A had won, he would have been happy
			\item Future Contrafactual == Future Subj:  S\=i A vincat, sit beatus:  If A should win, he would be happy
		\end{enumerate}
	\end{enumerate}
\end{enumerate}}

\newpage

\latline
{Pr\={\macron o}g\-en\-i\=em s\-ed \-en\=im Tr\=oj\={\macron a}n\={\macron o} \={\macron a} s\=ang\-{ui}ne d\={\macron u}c\={\macron i}}
{But indeed she had once heard that the progeny derived of Trojan blood}
{110}
{She had heard that a race from a Trojan bloodline was being led}
{\begin{enumerate}
	\item The infinitive should match the tense of the main verb
\end{enumerate}}

\latline
{\={au}d\-i\-er\=at T\-yr\-i\={\macron a}s \={\macron o}l\=im qu{\ae} v\=ert\-er\-et \=arc\={\macron e}s;}
{would overturn the fortresses of the Tyrians}
{150}
{which was of the characteristic to overturn the Tyrian citadels}
{\begin{enumerate}
	\item Why the subjunctive for ``v\=ert\-er\-et?''  It's a \textbf{purpose clause}, App 359.  ``vincit ut homines duceret''
	\item Relative clauses have two functions
	\begin{enumerate}
		\item Replace ``ut''
		\item \textbf{Relative clause of characteristic}
	\end{enumerate}
\end{enumerate}}

\latline
{h\=inc p\-op\-ul\=um l\={\macron a}t\={\macron e} r\={\macron e}g\=em b\=ell\={\macron o}qu\-e s\-up\=erb\=um}
{and that this origin a wide-ruling power, dominating in war,}
{110}
{she head that this race, widely ruling, proud in war, from here}
{\begin{enumerate}
	\item Re-insert ``she heard that'', ``audierat'' is still in effect
	\item Due to indirect discourse, ``populum'' is in the accusative; subject accusative
\end{enumerate}}

\newpage 

\latline
{v\=ent\={\macron u}r\sout{um }\=exc\-id\-i\={\macron o} L\-ib\-y\={\ae}: s\={\macron{\i}}c v\=olv\-er\-e P\=arc\={\macron a}s.}
{would fell Libya in war: these things had ben spun out by the Fates.}
{110}
{would venture to fell Lybia; she had heard the Fates spun thus}
{\begin{enumerate}
	\item excidi\={\macron o} Liby{\ae}, \textbf{dative of purpose}, App. 303
\end{enumerate}}

\latline
{\=Id m\-et\-u\={\macron e}ns v\-et\-er\=isqu\-e m\-em\=or S\={\macron a}t\=urn\-i\-a b\=ell\={\macron{\i}},}
{The daughter of Saturn, the chieftainess, fearing this prophecy and the mindful in the Trojan war}
{110}
{Saturnia, fearing it, and mindful of the ancient war}
{}

\latline
{pr\={\macron{\i}}m\-a qu\-od \=ad Tr\=oj\=am pr\={\macron o} c\={\macron a}r\={\macron{\i}}s g\=ess\-er\-at \=Arg\={\macron{\i}}s}
{thus in favor of her dear Argives to guided them to Troy}
{110}
{the foremost female diety, waged, on behalf of the Argives, against Troy}
{}

\newpage

\latline
{(n\=ecd\sout{um }\-et\-i\=am c\={au}s\sout{{{\ae}} }\={\macron{\i}}r{\macron a}rum s\={{\ae}}v\={\macron{\i}}qu\-e d\-ol\={\macron o}r\={\macron e}s}
{( for her reasons for savage sorrow and wrath }
{110}
{}
{}

\latline
{\=exc\-id\-er\=ant \-an\-im\={\macron o}; m\-an\-et \=alt\={\macron a} m\=ent\-e r\-esp\=ot\=um }
{had not yet calmed; she held, deep in her mind, the response,}
{110}
{she held deeply in her mind: the stored }
{
\begin{enumerate}
	\item ``anim\=o'' is ablative of separation
\end{enumerate}
}

\latline
{j\={\macron u}d\-ic\-i\=um P\-ar\-id\=is spr\={\macron e}t\-aqu\sout{e }\-inj\={\macron u}r\-i\-a f\=orm\={\ae}}
{(1) the judgement of Paris that insulted her beauty, }
{110}
{judgment of Paris and the injury of her spurned beauty}
{}

\newpage

\latline
{\=et g\-en\-us \=inv\={\macron{\i}}s\sout{um }\=et  r\=apt\={\macron{\i}} G\-an\-ym\={\macron e}d\-is h\-on\={\macron o}r\={\macron e}s) ---}
{(2) the hated race and (3) the honors of snatched-away Ganymede )}
{110}
{and the hated race, and the honors of snatched Ganymede}
{}

\latline
{h\={\macron{\i}}s \=acc\={\macron e}ns\-a s\-up\=er j\=act\={\macron a}t\={\macron o}s \={\ae}qu\-or\-e t\={\macron o}t\={\macron o}}
{and long restrained, on the watery wastes, on account of these vexations, }
{110}
{Saturnia, enraged by these things, was also keeping the whole-sea-tossed-about}
{}

\latline
{Tr\={\macron o}\=as, r\=el\-iqu\-i\={\macron a}s D\-an\-a\=um \=atq\sout{ue }\=imm\={\macron{\i}}t\-is \-Ach\=ill\={\macron{\i}},}
{those Trojans left by the Dana\"ans and fierce Achilles }
{110}
{Trojans, spared by the Dana\"ans and looming Achilles}
{}

\newpage

\latline
{\=arc\={\macron e}b\=at l\=ong\={\macron e} L\-at\-i\={\macron o}, m\=ult\={\macron o}squ\-e p\-er \=ann\={\macron o}s}
{she held them continually far away from Latium, for many years }
{110}
{away from Latium at length, for many years}
{}

\latline
{\=err\={\macron a}b\=ant \={\macron a}ct\={\macron{\i}} f\={\macron a}t\={\macron{\i}}s m\-ar\-i\sout{a }\=omn\-i\-a c\=irc\=um.}
{they wandered, as necessitated by fates, circling all the seas.}
{110}
{they wandered, led by fate around the entire sea.}
{}

\latline
{[33] T\=ant\={\ae} m\={\macron o}l\-is \-er\=at R\={\macron o}m\={\macron a}n\=am c\=ond\-er\-e g\=ent\=em.}
{So many were the challenges to the founding of the Roman race.}
{110}
{So great of a burden was it to fund the Roman nation.}
{
\begin{enumerate}
	\item Tant{\ae} is a genitive of quality
	\item ``gens'' means family, or nation.  Caesar Iuli\={\i}:  Julius Caesar
\end{enumerate}
}
