
\section{Virgil's {\AE}neid:  34-39} % (fold)

\latline
  {[34] V\=ix \={\macron e} c\={\macron o}nsp\=ect\={\macron u} S\-ic\-ul\={\ae} t\=ell\={\macron u}r\-is \-in \=alt\=um }
  {Barely from sight of the lands of Sicily, on the deep, the happy men}
  {110} %
  {Corrected}
  {
	\begin{enumerate}
		\item ``in altum'':  in the deep, the ocean
	\end{enumerate}
  }

\latline
  {v\={\macron e}l\-a d\-ab\=ant l\={\ae}t\={\macron{\i}}\sout{ e}t sp\={\macron u}m\={\macron a}s s\-al\-is \={\ae}r\-e r\-u\={\macron e}b\=ant,}
  {let free the sails and the bronze prows rushed into the salty waves,  }
  {110} %
  {Corrected translation}
  {
	\begin{enumerate}
		\item ``vela d\={a}re'' :  \emph{idiom.} ``to set sail''
		\item sp\={u}m\={a}s salis:  ``foam of the salt.''  here ``salt'' is functioning metonymously
	\end{enumerate}
  }

\latline
  {c\=um J\={\macron u}n\sout{o }\={\ae}t\=ern\=um s\=erv\={\macron a}ns s\=ub p\=ect\-or\-e v\=uln\={\macron u}s}
  {while Juno, eternally yoked under by her wounded pride}
  {110} %
  {when Juno, preserving eternally the wound in her breast}
  {
  }


\latline
  {h\={\ae}c s\={\macron e}c\=um:  ``M\={\macron e}n\sout{e }\=inc\=ept\={\macron o} d\={\macron e}s\=ist\-er\-e v\=ict\=am}
  {said these things to herself:  ``Am I, defeated, to desist---a victim--}
  {180} %
  {said these things:  Who am I, beatend to desist from my undertakings}
  {
	\begin{enumerate}
		\item Accusative of exclamation cf. Acc 318
		\item ``that I decease from my undertakings''
		\item \={\i}nc\={e}pt\={o}:  ablative of separation
		\item \={I}tali\={a}:  ablative of place from which
		\item Review of leading questions
		\item \begin{enumerate}
			\item -ne, expects yes or no
			\item -nonne, expects yes
			\item -num, expects no
		\end{enumerate}
	\end{enumerate}
  }

\latline
  {n\=ec p\=oss\sout{e }\={\macron{I}}t\-al\-i\={\macron a} T\-e\-ucr\={\macron o}r\sout{um }\={\macron a}v\=ert\-er\-e r\={\macron e}g\=em?}
  {or am i unable to avert the rule of Italy by the Trojans? }
  {110} %
  {nor am I unable to turn away the King of Trojans from Italy?}
  {
  }

\latline
  {Q\={ui}pp\-e v\-et\=or f\={\macron a}t\={\macron{\i}}s.  P\=all\=asn\sout{e }\=ex\={\macron u}r\-er\-e cl\=ass\=em}
  {Indeed I am forbidden from this by the fates.  But did not Pallas burn up the Greek fleet }
  {110} %
  {Corrected translation}
  {
  }



\latline
  {\=Arg\={\macron{\i}}v\sout{um }\=atqu\sout{e }\=ips\={\macron o}s p\-ot\-{u}\=it s\=umm\=erg\-er\-e p\=ont\={\macron o}}
  {and also wasn't she able to drown them in the sea}
  {110} %
  {Corrected translation}
  {
  }

\latline
  {\={\macron u}n\-i\-us \=ob n\=ox\sout{am }\=et f\=ur\=i\={\macron a}s \=Aj\={\macron a}c\=is \-O\-il\={\macron{e}\macron{\i}}}
  {on account of the crime and furies of one Ajax, son of Oliei?}
  {110} %
  {Corrected translation}
  {
  }

\latline
  {\=Ips\-a J\-ov\=is r\-ap\-id\=um j\-ac\-ul\={\macron a}t\sout{a }\={\macron e} n\={\macron u}b\-ib\-us \=ign\=em}
  {She herself, by throwing the bolts of Jupiter swiftly from the clouds}
  {110} %
  {She herself hurled from the clouds the swift bolts of Jove}
  {
  }

\latline
  {d\=isj\={\macron e}c\=itqu\-e r\-at\={\macron e}s \={\macron e}v\=ert\-itqu\-e \={\ae}qu\-or\-a v\=ent\={\macron{\i}}s,}
  {dispersed the ships and overturned them with roused waves and also snatched up }
  {110} %
  {and scattered the ships and overturned the sea with winds,}
  {
  }

\latline
  {\=ill\=u\sout{m e}xsp\={\macron{\i}}r\=ant\=em tr\={\macron a}nsf\={\macron{\i}}x\={\macron o} p\=ect\-or\-e fl\=amm\={\macron a}s}
  {and snatched up that single exhaling man with a burning chest in a}
  {110} %
  {Corrected translation}
  {
  }

\latline
  {t\=urb\-in\-e c\=orr\-ip\-u\=it sc\-op\-ul\={\macron o}qu\sout{e }\={\macron{\i}}nf\={\macron{\i}}x\-it \-ac\={\macron u}t\={\macron o};}
  {whirlwind  and impaled him on a sharp rock formation;}
  {110} %
  {Corrected translation}
  {
  }

\latline
  {\=ast \-eg\-o, qu\={\ae} d\={\macron{\i}}v\sout{um }\=inc\={\macron e}d\={\macron o} r\={\macron e}g\={\macron{\i}}n\-a J\-ov\=isqu\=e}
  {Certainly at least I, who stride as queen of the Gods and am }
  {110} %
  {Corrected translation}
  {
  }

\latline
  {\=et s\-or\-or \=et c\=onj\={\macron u}nx, \={\macron u}n\={\macron a} c\=um g\=ent\-e t\-ot \=ann\={\macron o}s}
  {both sister and spouse of Jove, shall do something with respect to those I }
  {110} %
  {Corrected translation}
  {
	\begin{enumerate}
		\item ablative + ``gente''
	\end{enumerate}
  }

\latline
  {b\=ell\-a g\-er\sout{o.  }\=Et q\={ui}squ\=am n\={\macron u}m\=en J\={\macron u}n\={\macron o}n\-is \-ad\={\macron o}r\=at}
  {fought for so many years.  And would anyone adore the power of Juno}
  {110} %
  {Corrected translation}
  {
  }

\latline
  {[49] pr\={\ae}t\=er\sout{\macron a }\={au}t s\=uppl\=ex \={\macron a}r\={\macron{\i}}s \=imp\={\macron o}n\-et h\-on\={\macron o}r\=em?}
  {besides or offer meekly as a supplicant at altars for her honor?''}
  {110} %
  {Corrected translation}
  {
  }

