f\section{Vergil's {\AE}neid: 201-250} % (fold)

\latline
  {L\={\macron {a}}oco\={\macron {o}}n, ductus Nept\={\macron {u}}n\={\macron {o}} sorte sacerd\={\macron {o}}s,}
  { Solemn Laoco\"on, a priest having been drawn unto Neptune, }
  {110}
  { Laoco\"on, a priest chosen by lot to Neptune, }
  { Notes }


\latline
  {sollemn\={\macron {\i}}s taurum ingentem mact\={\macron {a}}bat ad \={\macron {a}}r\={\macron {a}}s.}       
	{was slaughtering a gigantic bull on the altars  }
  {110}
  {was slaughtering a gigantic bull at the solemn altars.  }
  { Notes }


\latline
  {Ecce autem gemin\={\macron {\i}} \={\macron {a}} Tened\={\macron {o}} tranquila per alta}
	{ Moreover behold twin snakes with gigantic eyes, through tranquil water  }
  {110}
  { Behold, however, twin snakes with gigantic coils, through tranquil water  }                
  { Notes }


\latline
  {(horr\={\macron {e}}sc\={\macron {o}} ref\={\macron {e}}rens) imm\={\macron {e}}ns\={\macron {\i}}s orbibus angu\={\macron {e}}s}
	{ from Tenedos (horrible to recall!) }
  {110}
  { from Tenedos (I shudder recalling this!) }
  { Notes }


\latline
  {[\textbf{205}] incumbunt pelag\={\macron {o}} pariterque ad l\={\macron {\i}}tora tendunt;}
	{ fell upon the sea and wound, side by side, to the shore;  }
  {110}
  { fall upon the sea and wind, side by side, to the shore;  }   
  { Notes }


\latline
  {pectora qu\={\macron {o}}rum inter fl\={\macron {u}}ct\={\macron {u}}s arr\={\macron {e}}cta jub{\ae}que}
	{ the breasts of which, between waves, surfaced up  }
  {110}
  { whose breasts, between waves, having been lifted up between the waves,  }
  { Notes }


\latline
  {sanguine{\ae} supertant und\={\macron {a}}s; pars c\={\macron {e}}tera pontum}
	{ and the bloody crests of which surmounted the wave; the rest lay on  }
  {110}
  { and the bloody crests of which surmount the waves; the rest lays upon  }
  { Notes }


\latline
  {p\={\macron {o}}ne legit sinuatque im\={\macron {e}}nsa vol\={\macron {u}}mine terga.}
	{ the sea and slithers behind with immense coils. }
  {110}
  { the sea from behind and twists their immense back sides with coils. }   
  { Notes }


\latline
  {Fit sonitus sp\={\macron {u}}mante sal\={\macron {o}}; jamque arva ten\={\macron {e}}bant }
	{ There was a sound, with salt having frothed; and now already they were seeking the fields }
  {110}
  { A sound is made, with salt frothing; and now already they were holding the fields }
  { Notes }


\latline
  {[\textbf{210}] ardent\={\macron {\i}}sque ocul\={\macron {o}}s suffect\={\macron {\i}} sanguine et ign\={\macron {\i}}}
	{ and with their eyes having been supplied with burning blood and having burned  }
  {110}
  { and they, having supplied their eyes burning eyes with blood and fire  }
  { Notes }


\latline
  {s\={\macron {\i}}bila lamb\={\macron {e}}bant lingu\={\macron {\i}}s vibrantibus \={\macron {o}}ra.}
  { they lick hissing mouths  }
  {110}
  { they were licking their hissing mouths with darting tongues.  }
  { Notes }


\latline
  {Diffugimus v\={\macron {\i}}s\={\macron {u}} exsangu\={\macron {e}}s.  Ill\={\macron {\i}} agmine cert\={\macron {o}}}
	{ Drained of blood we flee this sight.  Those men, with a brave column }
  {110}
  { Drained of blood from sight we flee this.  Those snakes, in a straight line }
  { Notes }


\latline
  {L\={\macron {a}}oco\"{o}nta petunt; et pr\={\macron {\i}}mum parva du\={\macron {o}}rum}
	{ seek Laoco\"on; and first an embracing snake entwines the  }
  {110}
  { seek Laoco\"on; and first each snake having embraced the little   }   
  { Notes }


\latline
  {corpora n\={\macron {a}}t\={\macron {o}}rum serp\={\macron {e}}ns amplexus uterque}
	{ each of little bodies of his two sons }
  {110}
  { bodies of his two sons }                                    
  { Notes }


\latline
  {[\textbf{215}] implicat et miser\={\macron {o}}s mors\={\macron {u}} d\={\macron {e}}pascitur art\={\macron {u}}s:}
	{ and they devour the miserable sons with bites to their joints; }
  {110}
	{ they entwine them and devour the miserable joints (limbs) with biting; }
  { Notes }


\latline
  {p\=ost \=ips\sout{um }\={au}x\-il\-i\={\macron {o}} s\-ub\-e\=unt\sout{em }\=ac t\={\macron {e}}l\-a f\-er\=ent\=em}
	{ thereafter they seized he himself as he was going down to help and carrying a spear }
  {110}
	{ thereafter they seize Laco\"on himself as he was going down to help and carrying weapons }
  { Notes }


\latline
  {c\=orr\-ip\-i\=unt sp\={\macron {\i}}r\={\macron {\i}}squ\-e l\-ig\=ant \=ing\=ent\-ib\-us; \=et j\=am}
  { and bind him in immense coils; and now  }
  {110}
	{ttt}
  { Notes }


\latline
  {b\=is m\-ed\-i\sout{um }\=ampl\=ex\={\macron {\i}}, b\-is c\-oll\={\macron {o}} squ\={\macron {a}}m\-e\-a c\=irc\=um}
  { twice in the middle is embraced, with scaly backsides twice having been placed around the neck   }
  {110}
  { having twice enfolded him around the middle, with scaly backsides twice having been placed around his neck }
  { \begin{enumerate}
  	\item having enfolded him around the middle twice
  	\item When you have a perfect active participle derived from a deponent verb, you should drop the ``been'' from ``having been''
  \end{enumerate} }


\latline
  {t\=erg\-a d\-at\={\macron {\i}} s\-up\-er\=ant c\-ap\-it\sout{e }\=et c\=erv\={\macron {\i}}c\-ib\-us \=alt\={\macron {\i}}s.}
  { they overcome with heads and necks high. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{220}] \=Ill\-e s\-im\=ul m\-an\-ib\=us t\=end\=it d\={\macron {\i}}v\=ell\-er\-e n\={\macron {o}}d\={\macron {o}}s}
  { At the same time, that man tried with his hands to tear apart the knots, }
  {190}
  { At the same time, that man, having soaked the band in blood and dark poison struggles with hands to tear apart the knots, }
  { \begin{enumerate}
  	\item This is the ``middle voice'' again. Take perfund\={o}/perfundere/perfud\={\i}/perf\={u}sus. Take perf\={u}sus: perfect,
    passive participle ``having been soaked.'' In latin we do not have perfect active participles, therefore it is hard for verbs
    to take a direct object. Sometimes you want to use a perfect sense, but have it take a direct object. Most verbs cannot take a
    DO when they are in the perfect. Stealing from Greek, it keeps the perfect, but allows it to take a direct object; thus
    ``having soaked.'' The sense ``having \emph{x}-ed for himself'' then ``having \emph{been }\emph{x}-ed with respect to.'' This
    sounds weird. Thus make it ``having soaked the bands.''
		\item manus...juvenem revinctum :  ``the youth having been tied up in respect his hands''
		\item havinrg had something done to you, or having done something to yourself
  \end{enumerate} }


\latline
  {p\=erf\={\macron {u}}s\=us s\-an\-i\={\macron {e}} v\=itt\={\macron {a}}s \={\macron {a}}tr\={\macron {o}}qu\-e v\-en\={\macron {e}}n\={\macron {o}},}
  { his garlands stained with gore and black venom, }
  {110}
  { \emph{v.s.} }
  { Notes }


\latline
  {cl\={\macron {a}}m\={\macron {o}}r\={\macron {e}}s simul horrend\={\macron {o}}s ad s\={\macron {\i}}dera tollit:}
  { all the while he carried horrific cries to the stars: }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {qu\={\macron {a}}lis m\={\macron {u}}g\={\macron {\i}}tus, f\={\macron {u}}git cum saucius \={\macron {a}}ram}
  { a bellow such as, when a wounded bull flees the altar }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {taurus et incertam excussit cerv\={\macron {\i}}ce sec\={\macron {u}}rim.}
  { and shook off an uncertain axe from his neck.  }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{225}] At gemin\={\macron {\i}} l\={\macron {a}}ps\={\macron {u}} d\={\macron {e}}l\={\macron {u}}bra ad summa drac\={\macron {o}}n\={\macron {e}}s}
  { Furthermore the twin dragons fly away in a gliding motion toward the the highest temple }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {effugiunt s{\ae}v{\ae}que petunt Tr\={\macron {\i}}t\={\macron {o}}nidis arcem,}
  { and both go in the citadel of savage Minerva, }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {sub pedibusque de{\ae} clipe\={\macron {\i}}que sub orbe teguntur.}
  { under the feet of the goddess and hind themselves behind her shield. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {Tum v\={\macron {e}}r\={\macron {o}} tremefacto novus per pectora c\={\macron {u}}nct\={\macron {\i}}s}
  { Then indeed a new fear winds through the terrified breasts of all }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {\={\macron {\i}}nsinuat pavor, et scelus expendisse merentem}
  { and they hold that Laoco\"on expiated his merited crime  }
  {110}
  { and they hold that deserving Laoco\"on expiated his crime  }
  { Notes }


\latline
  {[\textbf{230}] L\={\macron {a}}oco\"{o}nta ferunt, sacrum qu\={\macron {\i}} cuspide r\={\macron {o}}bur }
  { he who had offended the sacred oak with his point }
  {110}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item Since we are using perfect subjunctive; these are relative clauses of characteristic
  \end{enumerate} }


\latline
  {l{\ae}serit et terg\={\macron {o}} sceler\={\macron {a}}tam intorserit hastam.}
  { and he who threw his wicked spear into its flank. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {D\={\macron {u}}cendum ad s\={\macron {e}}d\={\macron {e}}s simul\={\macron {a}}crum \={\macron {o}}randaque d\={\macron {\i}}v{\ae}}
  { They clamor that the likeness must be lead towards our homes and that the power of the goddess  }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {n\={\macron {u}}mina concl\={\macron {a}}mant.}
  { must be sought. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {D\={\macron {\i}}vidimus m\={\macron {u}}r\={\macron {o}}s et moenia pandimus urbis.}
  { We open our ramparts and spread open the walls of the city. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{235}] Accingunt omn\={\macron {e}}s oper\={\macron {\i}} pedibusque rot\={\macron {a}}rum}
  { All men gird for the work and and place gliding wheels under the feet  }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {subjiciunt l\={\macron {a}}ps\={\macron {u}}s, et stuppea vincula coll\={\macron {o}}}
  { and extend hempen lines from the neck: }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {intendunt: scandit f\={\macron {a}}t\={\macron {a}}lis m\={\macron {a}}china m\={\macron {u}}r\={\macron {o}}s}
  { the fatal machine mounts the walls  }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {f\={\macron {e}}ta arm\={\macron {\i}}s.  Puer\={\macron {\i}} circum inn\={\macron {u}}pt{\ae}que puell{\ae}}
  { pregnant with weapons.  All around boys and maidens }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {sacra canunt f\={\macron {u}}nemque man\={\macron {u}} contingere gaudent:}
  { sing sacred songs and are glad to touch the line with their hand: }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{240}] illa subit medi{\ae}que min\={\macron {a}}ns inl\={\macron {a}}bitur urb\={\macron {\i}}.}
  { that thing enters and, towering, glides into the center of the city. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {\={\macron {O}} patria, \={\macron {O}} d\={\macron {\i}}vum domus \={\macron {I}}lium et incluta bell\={\macron {o}}}
  { O my homeland, O Illium home of the gods and famous walls renowned in war }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {moenia Dardanidum! quater ips\={\macron {o}} in l\={\macron {\i}}mine port{\ae}}
  { of Dardanium!  four times the thing stumbled on the very threshold of the gates }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {substitit atque uter\={\macron {o}} sonitum quater arma ded\={\macron {e}}re;}
  { and also four times from the womb the arms gave a crash; }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {\={\macron {\i}}nst\={\macron {a}}mustamen immemor\={\macron {e}}s c{\ae}c\={\macron {\i}}que fur\={\macron {o}}re}
  { nevertheless we heedless people, blind with zeal press on and }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{245}] et m\={\macron {o}}nstrum \={\macron {\i}}nf\={\macron {e}}l\={\macron {\i}}x sacr\={\macron {a}}t\={\macron {a}} sistimus arce.}
  { we even set the unhappy idol in the sacred citadel. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {Tunc etiam f\={\macron {a}}t\={\macron {\i}}s aperit Cassandra fut\={\macron {u}}r\={\macron {\i}}s}
  { Then even Cassandra appears to speak the fates of things to come }
  {110}
  { Cassandra, who is never believed by the Teucians by the order of the god, then even opens }
  { Notes }


\latline
  {\={\macron {o}}ra de\={\macron {\i}}  juss\={\macron {u}} n\={\macron {o}}n umquam cr\={\macron {e}}dita Teucr\={\macron {\i}}s.}
  { mouth of the gods that will never be believed by Teucrians.  }
  {110}
  { her mouth to foretell the fates.  }
  { Notes }


\latline
  {N\={\macron {o}}s d\={\macron {e}}l\={\macron {u}}bra deum miser\={\macron {\i}}, quibus ultimus esset}
  { We miserable men, for whom that day would be the last, decked the sanctuaries of the gods }
  {110}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item Relative subjunctive, with adversative force
  \end{enumerate} }


\latline
  {ille di\={\macron {e}}s, f\={\macron {e}}st\={\macron {a}} v\={\macron {e}}l\={\macron {a}}mus fronde per urbem.}
  { with festal branches throughout the city. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{250}] Vertitur intere\={\macron {a}} c{\ae}lum et ruit \={\macron {O}}cean\={\macron {o}} nox}
  { Meanwhile heaven is turned and night rushes from Oceanus }
  {110}
  { CorrectedTranslation }
  { Notes }
