%
%  untitled
%
%  Created by Steven Harms on 2009-02-03.
%  Copyright (c) 2009 Steven G. Harms. All rights reserved.
%
\documentclass[]{book}

% Use utf-8 encoding for foreign characters
\usepackage[utf8]{inputenc}

% Setup for fullpage use
\usepackage{fullpage}

% Uncomment some of the following if you use the features
%
% Running Headers and footers
%\usepackage{fancyhdr}


% Multipart figures
%\usepackage{subfigure}

% More symbols
%\usepackage{amsmath}
%\usepackage{amssymb}
%\usepackage{latexsym}

% Surround parts of graphics with box
\usepackage{boxedminipage}

% Package for including code in the document
\usepackage{listings}

% If you want to generate a toc for each chapter (use with book)
\usepackage{minitoc}

% This is now the recommended way for checking for PDFLaTeX:
\usepackage{ifpdf}

\usepackage{setspace}
\usepackage{ulem}
\usepackage{framed}
\usepackage[en]{metre}
\usepackage{latinpoetry}
\usepackage{wasysym}

\ifpdf
\usepackage[pdftex]{graphicx}
\else
\usepackage{graphicx}
\fi
\title{Vergil's {\AE}neid}
\author{ Translated By:  Steven G. Harms }

\date{2009-02-03}

\begin{document}
\Elevatio{2}

\ifpdf
\DeclareGraphicsExtensions{.pdf, .jpg, .tif}
\else
\DeclareGraphicsExtensions{.eps, .jpg}
\fi

% Set the enum to NOT look like an outline, defined in latinpoetry.sty
\enumstyle 

\maketitle
\tableofcontents

\chapter{Chapter 1}
%\input{aeneid_1_1-33}
%\input{aeneid_1_34-49}
%\input{aeneid_1_50-88}
%\input{aeneid_1_89-117}
%\input{aeneid_1_118-141}
%\input{aeneid_1_142-173}


\latline
  {Ac pr\={\macron {\i}}mum silic\={\macron {\i}} scintillam exc\={\macron {u}}dit Ach\={\macron {a}}t\={\macron {e}}s}
  { And then Achates strikes flint to spark }
  {110}
  { First Achates struck out a spark with a flint }
  { Notes }



\latline
  {susc\={\macron {e}}pitque ignem foli\={\macron {\i}}s atque \={\macron {a}}rida circum}
  { and catches fire on leaves and gives it nearby }
  {110}
  { and caught up fire in the leaves and gives it nearby }
  { Notes }


\latline
  {n\={\macron {u}}tr\={\macron {\i}}menta dedit rapuitque in f\={\macron {o}}mite flamam.}
  { fuel and snatches up the flames in tinder. }
  {110}
  { fuel, and catches it up in the tinder. }
  { Notes }


\latline
  {Tum Cererem corruptam und\={\macron {\i}}s Cere\={\macron {a}}liaque arma}
  { Then they take out grain ruined by vawes, and grain tools, }
  {110}
  { They bring out the wave-corrupted grain and tools of Ceres }
  { Notes }



\latline
  {expediunt fess\={\macron {\i}} r\={\macron {e}}rum, fr\={\macron {u}}g\={\macron {e}}sque recept\={\macron {a}}s}
  { weary of all things and both parch the  }
  {110}
  { weary of all things and the salvaged grain }
  { Notes }


\latline
  {et torr\={\macron {e}}re parant flamm\={\macron {\i}}s et frangere sax\={\macron {o}}.}
  { rescued grain on the flames and grind it on rock. }
  {110}
  { and prepare to roast [it] with flames and break it with rock. }
  { Notes }


\latline
  {Aen\={\macron {e}}\={\macron {a}}s scopulum intere\={\macron {a}} c\={\macron {o}}nscendit, et omnem}
  { Meanwhile Aeneas climbs a rock, searches all views }
  {110}
  { Meanwhile {\AE}neas climbs a cliff and  }
  { Notes }



\latline
  {pr\={\macron {o}}spectum l\={\macron {a}}t\={\macron {e}} pelag\={\macron {o}} petit, Anthea s\={\macron {\i}} quem}
  { far and wide on the see, if he might see anything of wind-thrown }
  {110}
  { seeks a view on the entire see if he may  }
  { \begin{enumerate}
  	\item s\={\i} marks an indirect question
  	\item \textbf{quem} is operating as \textbf{aliquem}, a general supposition per the ``nisi / s\={\i}'' rules
  \end{enumerate} }


\latline
  {jact\={\macron {a}}tum vent\={\macron {o}} videat Phrygi\={\macron {a}}sque bir\={\macron {e}}m\={\macron {\i}}s}
  { Antheus and Phrygian galleys }
  {110}
  { see some wind-tossed Antheus, and if he may see the Phrygian triremes }
  { Notes }


\latline
  {aut Capyn aut cels\={\macron {\i}}s in puppibus arma Ca\={\macron {\i}}c\={\macron {\i}}.}
  { or either Capys or the shields of Caicus on towering decks. }
  {110}
  { CorrectedTranslation }
  { Notes }



\latline
  {N\={\macron {a}}vem in c\={\macron {o}}nspect\={\macron {u}} n\={\macron {u}}llam, tr\={\macron {\i}}s l\={\macron {\i}}tore cerv\={\macron {o}}s}
  { He sees no ships in view, but he sees three stags running on the  }
  {110}
  { No ship in sight, he sees three stags on the beach }
  { Notes }


\latline
  {pr\={\macron {o}}spicit errant\={\macron {\i}}s; h\={\macron {o}}s t\={\macron {o}}ta armenta sequuntur}
  { shore, those, the full herd follows behind, }
  {110}
  { wandering, those stags, the whole herd follows from behind,  }
  { Notes }


\latline
  {\={\macron {a}} terg\={\macron {o}} et longum per vall\={\macron {\i}}s pascitur agmen.}
  { distantly, across the valley and they graze as a troop }
  {110}
  { distantly, across the valley, and grazes }
  { Notes }



\latline
  {C\={\macron {o}}nstitit h\={\macron {\i}}c arcumque man\={\macron {u}} celer\={\macron {\i}}sque saggit\={\macron {a}}s}
  { Here Aeneas sets and snatches both bow and arrows with }
  {110}
  { He sets place here and grabbed with his hand the bow and swift arrows  }
  { \begin{enumerate}
  	\item C\={o}nstitit is \textbf{perfect}
  \end{enumerate} }


\latline
  {corripuit, f\={\macron {\i}}dus quae t\={\macron {e}}la ger\={\macron {e}}bat Ach\={\macron {a}}t\={\macron {e}}s,}
  { a swift hand, weapons which faithful Achates had brought  }
  {110}
  { which faithful Achates was carrying }
  { Notes }


\latline
  {duct\={\macron {o}}r\={\macron {e}}sque ips\={\macron {o}}s pr\={\macron {\i}}mum capita alta ferent\={\macron {\i}}s}
  { and he lays low those very leaders with branching horns, then he lays low the herd  }
  {110}
  { and he lays low first those leaders themselves, carrying their high heads with branching  }
  { Notes }



\latline
  {cornibus arbore\={\macron {\i}}s sternit, tum vulgus et omnem}
  { he then attacks the herd and  }
  {110}
  { horns and then mixes the entire herd }
  { Notes }


\latline
  {miscet ag\={\macron {e}}ns t\={\macron {e}}l\={\macron {\i}}s nemora inter frondea turbam;}
  { confuses all, sending the confusion into the leafy grove with his weapons }
  {110}
  { flushing them into the leafy grave with his weapons }
  { Notes }


\latline
  {nec prius absistit quam septem ingentia victor }
  { not stopping until the victor gathers seven huge }
  {110}
  { And he didn't stop before the victor slays }
  { \begin{enumerate}
  	\item prius{\ldots}quam is ``\textbf{priusquam}'' which is always separated
  \end{enumerate} }



\latline
  {corpora fundat hum\={\macron {\i}} et numerum cum n\={\macron {a}}vibus aequet.}
  { carcasses on the ground, he equals the count with the number of his ships. }
  {110}
  { 7 huge carcasses to the earth and he makes equal the number to the ships. }
  { Notes }


\latline
  {Hinc portum petit et soci\={\macron {o}}s part\={\macron {\i}}tur in omn\={\macron {\i}}s.}
  { He finds the port and divides them among all his compatriots. }
  {110}
  { Then he seeks the port and divides them among all his compatriots. }
  { Notes }


\latline
  {V\={\macron {\i}}na bonus quae deinde cad\={\macron {\i}}s oner\={\macron {a}}rat Acest\={\macron {e}}s}
  { Good Acestes unloads Sicilian wine jars on the beach }
  {110}
  { He divides wine which good Acestes had  }
  { \begin{enumerate}
  	\item oner\={a}rat is poetic ``oneraverat'', thus perfect
  \end{enumerate} }



\latline
  {l\={\macron {\i}}tore Tr\={\macron {\i}}nacri\={\macron {o}} dederatque abeuntibus h\={\macron {e}}r\={\macron {o}}s}
  { and gives to departed heroes and [then] divided, and calmed their grieving hearts with words. }
  {110}
  { given on the Sicilian shore to the departing heroes }
  { Notes }


\latline
  {d\={\macron {\i}}vidit, et dict\={\macron {\i}}s maerentia pectora mulcet:}
  { Translation }
  {110}
  { and calms the gloomy hearts with these words }
  { Notes }


\latline
  {``\={\macron {O}} soci\={\macron {\i}} (neque enim ign\={\macron {a}}r\={\macron {\i}} sumus ante mal\={\macron {o}}rum),}
  { Translation }
  {110}
  { O my friends (O men who are not ignorant of more grave ills) }
  { Notes }


\latline
  {\={\macron {O}} pass\={\macron {\i}} gravi\={\macron {o}}ra, dabit deus h\={\macron {\i}}s quoque f\={\macron {\i}}nem.}
  { Translation }
  {110}
  { O suffering-greatly men, God will also give end to these things }
  { Notes }


\latline
  {V\={\macron {O}}s et Scyllaeam rabiem penitusque sonant\={\macron {\i}}s}
  { Translation }
  {110}
  { Youwho approached both the madness of Scilla and within }
  { Notes }


\latline
  {accestis scopul\={\macron {o}}s, v\={\macron {o}}s et Cycl\={\macron {o}}pia saxa }
  { Translation }
  {110}
  { crashing rocks, you both endured the Cyclops' stones and }
  { Notes }


\latline
  {expert\={\macron {\i}}:  revoc\={\macron {a}}te anim\={\macron {o}}s maestumque tim\={\macron {o}}rem}
  { Translation }
  {110}
  { his rocks, call back your strength, dispel gloomy fear }
  { Notes }


\latline
  {mittite; forsan et haec \={\macron {o}}lim meminisse juv\={\macron {a}}bit.}
  { Translation }
  {110}
  { perhaps, someday, it will be pleasing to remember even these things some day }
  { \begin{enumerate}
  	\item The ``et'' is stronger than ``and'' since it is not, strictly, required.  
  \end{enumerate} }


\latline
  {Per vari\={\macron {o}}s c\={\macron {a}}s\={\macron {u}}s, per tot discr\={\macron {\i}}mina r\={\macron {e}}rum}
  { Translation }
  {110}
  { Through various misfrotunes, through so many crises of things and dangerous circumstances, }
  { Notes }


\latline
  {tendimus in Latinum, s\={\macron {e}}d\={\macron {e}}s ubi f\={\macron {a}}t\={\macron {a}} qui\={\macron {e}}t\={\macron {a}}s}
  { Translation }
  {110}
  { we stretch ourselves to Latinum where the fates promise quiet homes }
  { Notes }


\latline
  {ostendunt; ill\={\macron {\i}} f\={\macron {a}}s r\={\macron {e}}gina resurgere Trojae.}
  { Translation }
  {110}
  { there, it is right for the kingdom of Troy to resurge. }
  { Notes }


\latline
  {D\={\macron {u}}r\={\macron {a}}te, et v\={\macron {o}}smet r\={\macron {e}}bus serv\={\macron {a}}te secund\={\macron {\i}}s.''}
  { Translation }
  {110}
  { Endure, and preserve yourselves with joyful things. }
  { Notes }


\latline
  {T\={\macron {a}}lia v\={\macron {o}}ce refert c\={\macron {u}}r\={\macron {\i}}sque ingentibus aeger}
  { Translation }
  {110}
  { Such thing he says with his voice and sick with huge cares  }
  { Notes }


\latline
  {spem vult\={\macron {u}} simulat, premit altum corde dol\={\macron {o}}rem.}
  { Translation }
  {110}
  { he feigns hope in his face, and he presses the pain deep in his heart. }
  { Notes }


\latline
  {Ill\={\macron {\i}} s\={\macron {e}} praedae accingunt dapibusque fut\={\macron {u}}r\={\macron {\i}}s:}
  { Translation }
  {110}
  { Those men gird themselves for booty and the coming feasts }
  { Notes }


\latline
  {tergora d\={\macron {\i}}ripiunt cost\={\macron {\i}}s et v\={\macron {\i}}scera n\={\macron {u}}dant;}
  { Translation }
  {110}
  { they tear the hide from the ribs and strip the vitals }
  { Notes }


\latline
  {pars in frusta secant veribusque trementia figunt,}
  { Translation }
  {110}
  { they dry them in equal parts and pierce them with shaking spears. }
  { Notes }


\latline
  {l\={\macron {\i}}tore aena locant ali\={\macron {\i}} flamm\={\macron {a}}sque ministrant.}
  { They place a kettle on the shore and provide it other flame. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {Tum v\={\macron {\i}}ct\={\macron {u}} revocant v\={\macron {\i}}r\={\macron {\i}}s, f\={\macron {u}}s\={\macron {\i}}que per herbam}
  { Translation }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {implentur veteris Bacch\={\macron {\i}} pinguisque fer\={\macron {\i}}nae.}
  { Translation }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {Postquam ex\={\macron {e}}mpta fam\={\macron {e}}s epul\={\macron {\i}}s m\={\macron {e}}nsaeque rem\={\macron {o}}tae,}
  { Translation }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {\={\macron {a}}miss\={\macron {o}}s long\={\macron {o}} soci\={\macron {o}}s serm\={\macron {o}}ne requ\={\macron {\i}}runt,}
  { Translation }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {spemque metumque inter dubi\={\macron {\i}}, seu v\={\macron {\i}}vere cr\={\macron {e}}dant}
  { Translation }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {s\={\macron {\i}}ve extr\={\macron {e}}ma pat\={\macron {\i}} nec jam exaud\={\macron {\i}}re voc\={\macron {a}}t\={\macron {o}}s.}
  { Translation }
  {110}
  { CorrectedTranslation }
  { Notes }




%219



\end{document}
