\section{Vergil's {\AE}neid:  214-253} % (fold)

\latline
  {Tum v\={\macron {\i}}ct\={\macron {u}} revocant v\={\macron {\i}}r\={\macron {\i}}s, f\={\macron {u}}s\={\macron {\i}}que per herbam}
  { Then, they call back their strength with food and they, set among grass }
  {110}
  { Then, they call back their courage with food and they, poured out among the grass }
  { Notes }


\latline
  {[\textbf{215}]implentur veteris Bacch\={\macron {\i}} pinguisque fer\={\macron {\i}}nae.}
  { fill themselves with old wine and fat venison. }
  {110}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item implentur is a ``middgle'' Greek form, meaning ``to do something for oneself''
  \end{enumerate} }


\latline
  {Postquam ex\={\macron {e}}mpta fam\={\macron {e}}s epul\={\macron {\i}}s m\={\macron {e}}nsaeque rem\={\macron {o}}tae,}
  { After hungers were dispelled by the feast and the banquet was finished }
  {110}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item Occasionally poetry shortens ``postquam'' to ``post.'' Typically you
    will be expecting ``post'' + accusative. If you don't find the accusative,
    it's proabbly the poetic ``postquam.''
  \end{enumerate} }


\latline
  {\={\macron {a}}miss\={\macron {o}}s long\={\macron {o}} soci\={\macron {o}}s serm\={\macron {o}}ne requ\={\macron {\i}}runt,}
  { they sought again their lost comrades with long conversation }
  {110}
  { they seek again their lost comrades with long conversation }
  { Notes }


\latline
  {spemque metumque inter dubi\={\macron {\i}}, seu v\={\macron {\i}}vere cr\={\macron {e}}dant}
  { the doubting men, between both hope and fear, assert whether their friends live }
  {110}
  { they waver between hope and fear and doubt whether the called men}
  { \begin{enumerate}
  	\item Independent subjunctives
		\item \textbf{hortatory} :  ``let us lead'':  ``agamus''
		\item \textbf{jussive}:  ``let him lead'': ``agatus''
		\item \textbf{deliberative}:  ``should we lead'': ``agamus?''
		\item Note that we're in indirect discourse, which is subject accusative +
    an infinitive. The accusative, ``vocat\={o}s'' occurs at the end of the
    line and needs to be carried back into the enumeration of their doubts
  \end{enumerate} }


\latline
  {s\={\macron {\i}}ve extr\={\macron {e}}ma pat\={\macron {\i}} nec jam exaud\={\macron {\i}}re voc\={\macron {a}}t\={\macron {o}}s.}
  { or if they died and if they can still hear. }
  {110}
  { are living, if they are enduring the last things, and if they no longer can hear them. }
  { \begin{enumerate}
  	\item We are in indirect discourse:  we have a subject accusative + infinitive
  	\item non{\ldots}jam == ``no longer''
  \end{enumerate} }

\latline
  {[\textbf{220}] Praecipu\={\macron {e}} pius {\AE}n\={\macron {e}}\={\macron {a}}s nunc \={\macron {a}}cris Oront\={\macron {\i}},}
  { Pious {\AE}neas weeps now especially for keen and faithful Orontes }
  {110}
  { Pious {\AE}neas weeps now especially for the fate of keen Orontes }
  { Notes }


\latline
  {nunc Amyc\={\macron {\i}} c\={\macron {a}}sum gemit et cr\={\macron {u}}d\={\macron {e}}lia s\={\macron {e}}cum}
  { and now laments the fate of Amycus and, to himself laments }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {f\={\macron {a}}ta Lyc\={\macron {\i}} fortemque Gy\={\macron {a}}n fortemque Cloanthum.}
  { the cruel fates of the Lycians and both of strong Gyas and strong Cloanthus }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {Et jam f\={\macron {\i}}nis erat, cum Juppiter aethere summ\={\macron {o}}}
  { And while the end was happening, Jupiter, sitting in highest heaven }
  {110}
  { And now there was an end [{\ldots}to the talking\ldots], Jupiter, looking down from highest heaven }
  { Notes }


\latline
  {dispici\={\macron {e}}ns mare v\={\macron {e}}livolum terr\={\macron {a}}sque jacent\={\macron {\i}}s}
  { seeing the sea strewn with full sails and spread-out lands }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{225}] l\={\macron {\i}}toraque et l\={\macron {a}}t\={\macron {o}}s popul\={\macron {o}}s, s\={\macron {\i}}c vertice cael\={\macron {\i}}}
  { and shores and wide-spead people, so sat in the summit of heaven }
  {110}
  { and shores and wide-spead people, so stood in the summit of heaven }
  { Notes }


\latline
  {c\={\macron {o}}nstitit et Libyae d\={\macron {e}}fixit l\={\macron {u}}mina r\={\macron {e}}gn\={\macron {\i}}s.}
  { he sat and fixed his ruling eyes to Libya. }
  {110}
  { and fixed his ruling eyes on Libya. }
  { Notes }


\latline
  {Atque illum t\={\macron {a}}l\={\macron {\i}}s jacantem pectore c\={\macron {u}}r\={\macron {a}}s}
  { Also, Venus addressed that god, that god with a breast buffeted by so many cares }
  {190}
  { Venus addresses that god, tossing about such cares in breast  }
  { \begin{enumerate}
  	\item jacantem comes from jacto\={o}(1) meaning \emph{buffeted}. Thus from
    \textbf{jact\={a}ns/jactantis}. Juppiter is in the accusative (being
    addressed by Venus), thus the present, active participle modifies him,
    i.e. jactantem.
    \item jaci\={o}, jacere, jec\={\i}, jactum:  To throw
    \item jace\={o}, jac\={e}re, : to lie
    \item jact\={o}(1):  to buffet
  \end{enumerate} }


\latline
  {tr\={\macron {\i}}stior et lacrim\={\macron {\i}}s ocul\={\macron {o}}s suff\={\macron {u}}sa nitent\={\macron {\i}}s}
  { and spoke to him, more sad than usual, having filled up her shining eyes with tears: }
  {110}
  { and addressed him, more sad than usual, having filled up her eyes with tears }
  { \begin{enumerate}
  	\item Note the ``middle'' verb type construct:  ``filled up her own eyes''
  	\item Literally the phrase is and with tears her eyes Venus-suffused with bright
  	\item suffused with tears with respect to eyes ––- cf. \textbf{Accusative of respect}
  \end{enumerate} }


\latline
  {adloquitur Venus: ``\={\macron {O}} qu\={\macron {\i}} r\={\macron {e}}s hominumque deumque}
  { O you, who ruling things both of man and gods, }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{230}] aetern\={\macron {\i}}s regis imperi\={\macron {\i}}s et fulmine terr\={\macron {e}}s,}
  { who rules with eternal imperium and frightens with the thunderbolt }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {quid meus Aen\={\macron {e}}\={\macron {a}}s in t\={\macron {e}} committere tantum,}
  { what such great thing could my Aeneas commit against you, }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {quid Tr\={\macron {o}}es potu\={\macron {e}}re, quibus tot f\={\macron {u}}nera pass\={\macron {\i}}s}
  { what could the trojans[do], for whom, having endured so many deaths}
  {110}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item ``quibus'' is dative of reference
  \end{enumerate} }


\latline
  {c\={\macron {u}}nctus ob \={\macron {I}}taliam terr\={\macron {a}}rum clauditur orbis?}
  { the world is shut off on account of Italy? }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {Cert\={\macron {e}} hinc R\={\macron {o}}m\={\macron {a}}n\={\macron {o}}s \={\macron {o}}lim volventibus ann\={\macron {\i}}s,}
  { Certainly here you promised once, as years roll by, that in this place }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{235}] hinc fore duct\={\macron {o}}r\={\macron {e}}s, revoc\={\macron {a}}t\={\macron {o}} \={\macron {a}} sanguine Teucr\={\macron {\i}},}
  { there would be Romans, in the roll of time, that they would be leaders, from the recalled blood of Teucer }
  {110}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item What is ``hinc'' referring to?  These men, the Trojans.
  \end{enumerate} }


\latline
  {qu\={\macron {\i}} mare, qu\={\macron {\i}} terr\={\macron {a}}s omn\={\macron {\i}}s dici\={\macron {o}}ne ten\={\macron {e}}rent,}
  { Romans who would hold the sea, and all lands with their power. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {pollicitus.  Quae t\={\macron {e}}, genitor, sententia vertit?}
  { What thing, father, changed your plan? }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {H\={\macron {o}}c equidem occ\={\macron {a}}sum Trojae tr\={\macron {\i}}st\={\macron {\i}}sque ru\={\macron {\i}}n\={\macron {a}}s}
  { Surely with this fact I tried to comfort myself about the fate of troy }
  {110}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item ``H\={o}c'' means, these things
  \end{enumerate} }


\latline
  {s\={\macron {o}}l\={\macron {a}}bar f\={\macron {a}}t\={\macron {\i}}s contr\={\macron {a}}ria f\={\macron {a}}ta repend\={\macron {e}}ns;}
  { with this fate, to balance those opposing fates; }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{240}] nunc eadem fort\={\macron {u}}na vir\={\macron {o}}s tot c\={\macron {a}}sibus \={\macron {a}}ct\={\macron {o}}s}
  { Now the same cursed fate follows those men having been led by misfortunes. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {\={\macron {\i}}nsequitur.  Quem d\={\macron {a}}s f\={\macron {\i}}nem, r\={\macron {e}}x magne, lab\={\macron {o}}rum?}
  { You give, what end, Great King, of their toils? }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {Ant\={\macron {e}}nor potuit medi\={\macron {\i}}s \={\macron {e}}l\={\macron {a}}psus Ach\={\macron {\i}}v\={\macron {\i}}s}
  { Having escaped from the midst of the Greeks, Antenor could }
  {110}
  { Slipped-away Antenor, from the midst of the Greeks, was able  }
  { \begin{enumerate}
  	\item Ant\={e}nor goes with \={e}l\={a}psus:  ``Having-slipped-away Antenor'' is the nominative
  \end{enumerate} }


\latline
  {Illyric\={\macron {o}}s penetr\={\macron {a}}re sin\={\macron {u}}s atque intima t\={\macron {u}}tus}
  { enter the openings of the Illyrians, and also could enter within }
  {110}
  { to penetrate the channels of the Illyrians and also penetrate the inmost  }
  { Notes }


\latline
  {r\={\macron {e}}gna Liburn\={\macron {o}}rum et fontem super\={\macron {a}}re Tim\={\macron {a}}v\={\macron {\i}},}
  { the safe kingdom of the Liburnians, and could cross the fountain of Timaveus }
  {110}
  { realm of the Liburnians safely, and he was able to cross the Timavi }
  { Notes }


\latline
  {[\textbf{245}] unde per \={\macron {o}}ra novem vast\={\macron {o}} cum mumere montis}
  { at that place where the sea runs rushing through nine gaping mouths }
  {110}
  { from where, through 9 gaping mouths mouths with the murmur of the mountain, rushing,  }
  { Notes }


\latline
  {it mare pr\={\macron {o}}ruptum et pelag\={\macron {o}} premit arva sonat\={\macron {\i}}.}
  { and with its flood takes the plains. }
  {110}
  { the water goes and overwhelms the fields with sounding wave. }
  { Notes }


\latline
  {H\={\macron {\i}}c tamen ille urbem Patav\={\macron {\i}} s\={\macron {e}}d\={\macron {e}}sque loc\={\macron {a}}vit}
  { Here, nevertheless, that man set both a city and Patavian seats }
  {110}
  { There, nevertheless that man was able to found the city of Patavium, and set seats }
  { \begin{enumerate}
  	\item Patav\={\i} is a genitive of apposition
  \end{enumerate} }


\latline
  {Teucr\={\macron {o}}rum et gent\={\macron {\i}} n\={\macron {o}}men dedit armaque fixit}
  { of Troy, gave name to the clan and fixed his arms. }
  {110}
  { of Teucer and gave name to the people, and fixed the arms of Troy. }
  { Notes }


\latline
  {Tr\={\macron {o}}ia, nunc placid\={\macron {a}} compostus p\={\macron {a}}ce qui\={\macron {e}}scit:}
  { Troy now sits still, having been set by calm peace  }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{250}] n\={\macron {o}}s, tua pr\={\macron {o}}geni\={\macron {e}}s, cael\={\macron {\i}} quibus adnuis arcem,}
  { we, your children, for whom you promise the citadel of heaven, we are set  }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {n\={\macron {a}}vibus (\={\macron {\i}}nfandum!) \={\macron {A}}miss\={\macron {\i}}s \={\macron {u}}n\={\macron {\i}}us ob \={\macron {\i}}ram}
  { with ships having been lost (for shame!) on account of the wrath of one,  }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {pr\={\macron {o}}dimur atque Ital\={\macron {\i}}s long\={\macron {e}} disjungimur \={\macron {o}}r\={\macron {\i}}s.}
  { we are being betrayed and also we are long separated from the borders of Italy. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {Hic piet\={\macron {a}}tis hon\={\macron {o}}s?  S\={\macron {\i}}c n\={\macron {o}}s in sc\={\macron {e}}ptra rep\={\macron {o}}nis?''}
  { Is this the honor of piety?  Thusly do you return the scepter unto us? }
  {110}
  { CorrectedTranslation }
  { \begin{enumerate}
  	\item You might expect ``hic'', but it's attracted to the gender of hon\={o}s
  \end{enumerate} }