\section{Vergil's {\AE}neid: 151-200} % (fold)

\latline
  {Quidve petunt?  Qu{\ae} relligi\={\macron {o}}?  Aut qu{\ae} m\={\macron {a}}china bell\={\macron {\i}}?'}
  { Or what do they seek?  Of what sort of offering?  Or of what machinery of war?' }
  {110}
  { Or what do they seek?  Of what religion?  Or machine of war?' }
  { Notes }


\latline
  {D\={\macron {\i}}xerat.  Ille dol\={\macron {\i}}s \={\macron {\i}}nstr\={\macron {u}}ctus et arte Pelasg\={\macron {a}}}
  { It was told.  That man, having been instructed by lies and by Pelasgian art,  }
  {110}
  { He had spoken. That man, having been instructed with lies and by Pelasgian art,  }
  { Notes }


\latline
  {sustulit ex\={\macron {u}}t\={\macron {a}}s vincl\={\macron {\i}}s ad s\={\macron {\i}}dera palm\={\macron {a}}s:}
  { liberated from chains, lifts his freed palmas to the stars: }        
  {110}
  { lifted liberated-from-chains palms, to the stars: }
  { Notes }


\latline
  {`V\={\macron {o}}s {\ae}tern\={\macron {\i}} ign\={\macron {e}}s, et n\={\macron {o}}n viol\={\macron {a}}bile vestrum}
  { `You eternal stars, I call [you] and especially you inviolable  }
  {110}
  { `You eternal stars, I call [you to bear witness] and especially [upon] your inviolable  }
  { Notes }


\latline
  {[\textbf{155}] testor n\={\macron {u}}men,' ait, `v\={\macron {o}}s \={\macron {a}}r{\ae} \={\macron {e}}ns\={\macron {e}}sque nefand\={\macron {\i}},}
  { divine power,' he said, `you, unspeakable altars and knives,  }            
  {110}
  { divine power,' he said, `you, altars and accursed knives,  }
  { Notes }


\latline
  {qu\={\macron {o}}s f\={\macron {u}}g\={\macron {\i}}, vitt{\ae}que deum, qu\={\macron {a}}s hostia gess\={\macron {\i}}:}
  { which I escaped, and the garlands of the gods, for which I was lead as sacrifice:  }
  {110}
  { which I escaped, and the garlands of the gods, which I, a sacrificial victim, wore:  }
  { Notes }


\latline
  {f\={\macron {a}}s mihi Graj\={\macron {o}}rum sacr\={\macron {a}}ta resolvere j\={\macron {u}}ra,}
  { it is just for me, the oaths of the Greeks having been dedicated, to dissolve [them] }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {f\={\macron {a}}s \={\macron {o}}disse vir\={\macron {o}}s atque amnia ferre sub aur\={\macron {a}}s,}
  { it is just to hate the men and also to bring all things to light, }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {s\={\macron {\i}} qua tegunt; teneor patri{\ae} nec l\={\macron {e}}gibus \={\macron {u}}ll\={\macron {\i}}s.  }
  { if they hide them; I am held to neither my homeland nor by the laws of the others. }    
  {110}                                                                                     
  { if anything hides; I am not held by any laws of my fatherland. }
  { \begin{enumerate}
  	\item qua, yet again == ``aliqua''
  \end{enumerate} }


\latline
  {[\textbf{160}] T\={\macron {u}} modo pr\={\macron {o}}miss\={\macron {\i}}s mane\={\macron {a}}s serv\={\macron {a}}taque serv\={\macron {e}}s}
  { Only let you remain, things having been promised, and, preserve, things having been preserved,     }
  {110}
  { May you only remain in your promises, and preserve faith, you Troy, having been saved  }
  { Notes }


\latline
  {Troja fidem, s\={\macron {\i}} v\={\macron {e}}ra feram, s\={\macron {\i}} magna rependam.}
  {trust, Troy, if I shall bear truthfully, if I shall bear you great reward. }
  {110}
  { if I shall speak truthfully and pay you back so handsomely. }
  { \begin{enumerate}
  	\item Pay from hang, as in the balances are equal. (``rependam'')
  \end{enumerate} }


\latline
  {Omnis sp\={\macron {e}}s Danaum et coept\={\macron {\i}} f\={\macron {\i}}d\={\macron {u}}cia bell\={\macron {\i}}}
  { The hope of all the Dan\"aans and fidelity, war having been undertaken,  }    
  {110}
  { The hope of all the Dan\"aans and confidence of the war having been begun,  }
  { Notes }


\latline
  {Palladis auxili\={\macron {\i}}s semper stetit.  Impius ex qu\={\macron {o}}}
  { always rested on the help of Pallas.  During which time the impious Diomedes  }
  {110}
  { always rested on the help of Pallas.  From the time when the impious Diomedes  }
  { \begin{enumerate}
  	\item ex qu\={o}:  From the time
  \end{enumerate} }


\latline
  {T\={\macron {y}}d\={\macron {\i}}d\={\macron {e}}s sed enim scelerumque inventor Ulix\={\macron {e}}s}
  { but indeed even more the author of evils, Ulysses, }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {[\textbf{165}] f\={\macron {a}}t\={\macron {a}}le adgress\={\macron {\i}} sacr\={\macron {a}}t\={\macron {o}} \={\macron {a}}velllere templ\={\macron {o}}}
  { attacked to snatch away from the sacred temple }                
  {110}
  { undertook to snatch away from the sacred temple }
  { Notes }


\latline
  {Palladium c{\ae}s\={\macron {\i}}s summ{\ae} cust\={\macron {o}}dibus arcis}            
  { the Palladium of the height of the citadel, the guards having been slain,  }    
  {110}
  { the fateful Palladium, the guards of the top of the citadel having been slain,  }
  { Notes }


\latline
  {corripu\={\macron {e}}re sacram effigiem manibusque cruent\={\macron {\i}}s}    
  { they snatched the sacred image and with bloody hands  }      
  {110}
  { they snatched away the sacred image and with bloody hands  }
  { Notes }


\latline
  {virgine\={\macron {a}}s aus\={\macron {\i}} d\={\macron {\i}}v{\ae} contingere vitt\={\macron {a}}s:}
  { dared to touch the virginal garlands of the goddess:  }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {ex ill\={\macron {o}} fluere ac retr\={\macron {o}} subl\={\macron {a}}psa referr\={\macron {\i}}}
  { from that time the hopes of the Dana\"ans ebbed and, things having slipped, were carried }
  {110}
  { from that time, the hopes of the Dana\"ans ebbed and were carried backwards, the hope having slipped, }
  { \begin{enumerate}
  	\item fluere and referr\={\i} are historical infinitives.  Translate according to context, quickens the narrative.
  \end{enumerate} }


\latline
  {[\textbf{170}] sp\={\macron {e}}s Danaum, fr\={\macron {a}}ct{\ae} v\={\macron {\i}}r\={\macron {e}}s, \={\macron {a}}versa de{\ae} m\={\macron {e}}ns.}
  { backwards, their powers having been broken, the mind of the goddess having been turned. }       
  {110}
  { and their powers were broken, the mind of the goddess was turned. }             
  { Notes }


\latline
  {Nec dubi\={\macron {\i}}s ea signa dedit Tr\={\macron {\i}}t\={\macron {o}}nia m\={\macron {o}}nstr\={\macron {\i}}s.}
  { By no uncertain portents gave Minverva these signs. }
  {110}
  { As no uncertain portents, Minerva gave these signs. }
  { Notes }


\latline
  {Vix positum castr\={\macron {\i}}s simul\={\macron {a}}crum: ars\={\macron {e}}re corusc{\ae}}
  { The image scarcely having been placed in the camps: glittering flames burned }                 
  {110}
  { The image was scarcely  placed in the camps: flashing flames blazeed up  }
  { Notes }


\latline
  {l\={\macron {u}}minibus flamm{\ae} arr\={\macron {e}}ct\={\macron {\i}}s, salusque per art\={\macron {u}}s}       
  { in her opened eyes and salty sweat came forth out of the }           
  {110}
  { in her opened eyes and salty sweat went over the }
  { Notes }


\latline
  {s\={\macron {u}}dor iit, terque ipsa sol\={\macron {o}} (m\={\macron {\i}}r\={\macron {a}}bile dict\={\macron {u}})}
  { joints, thrice the goddess herself even (amazing to say) }         
  {110}
 { joints, and thrice the goddess herself even (amazing to say) }
  { Notes }


\latline
  {[\textbf{175}] \={\macron {e}}micuit parmamque fer\={\macron {e}}ns hastamque trementem.}
  { dashed forth carrying both her shield and gigantic spear. }
  {110}
  { dashed forth carrying both her shield and shaking spear. }
  { Notes }


\latline
  {Extempl\={\macron {o}} temptanda fug\={\macron {a}} canit {\ae}quora Calch\={\macron {a}}s,}
  { Suddenly Calchas sang that waves are to be attempted with flight, }      
  {110}
  { Suddenly Calchas prophesies that waves are must be attempted with flight, }
  { Notes }


\latline
  {nec posse Argolic\={\macron {\i}}s exscind\={\macron {\i}} Pergama t\={\macron {e}}l\={\macron {\i}}s}
  { that it was not possible to cut out the citadel of Troy by such Argives }
  {110}
  { that Pergamum is not possible to cut be destroyed by Argive weapons }
  { Notes }


\latline
  {\={\macron {o}}mnia n\={\macron {\i}} repetant Arg\={\macron {\i}}s n\={\macron {u}}menque red\={\macron {u}}cant}
  { unless they took new auspices from Greece and would bring back with them the divine favor }
  {110}
  { unless they seek new auspices from Greece and  bring back with them the divine favor }
  { Notes }


\latline
  {quod pelag\={\macron {o}} et curv\={\macron {\i}}s s\={\macron {e}}cum \={\macron {a}}v\={\macron {e}}x\={\macron {e}}re car\={\macron {\i}}n\={\macron {\i}}s.}
  { which they brought with them from Greece in curved hulls. }       
  {160}
  { which they brought with them on the ocean from Greece in curved hulls. } 
  { \begin{enumerate}
  	\item Canit is taking as indirect statmement ``posse''
    \item A subordinate clause inside of indirect discourse always takes the subjunctive per ``oblique speech''
    \item You have to use sequence of tenses to make the subjunctives agree with ``posse'' :: needn't tronslate as
subjunctives, feel free to use indicatives.  In the same way that indirect questions work.
  \end{enumerate} }


\latline
  {[\textbf{180}] Et nunc quod patri\={\macron {a}}s vent\={\macron {o}} peti\={\macron {e}}re Myc\={\macron {e}}n\={\macron {a}}s,}
  { And then they sought the homelands of Mycenae with the wind }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {arma de\={\macron {o}}sque parant comit\={\macron {e}}s pelag\={\macron {o}}que rem\={\macron {e}}ns\={\macron {o}}}
  { they prepare arms and gods comrades and, the sea having been re-crossed,  }     
  {110}
  { the comrades prepare arms, and gods, and, the sea having been re-crossed,  }
  { Notes }


\latline
  {impr\={\macron {o}}v\={\macron {\i}}s\={\macron {\i}} aderunt.  Ita d\={\macron {\i}}gerit \={\macron {o}}mina Calch\={\macron {a}}s.}
  { suddenly they were present.  Indeed, Calchas explained everything. }              
  {110}
  { they will be here unexpectedly.  Indeed, Calchas explains the omens. }
  { Notes }


\latline
  {Hanc pr\={\macron {o}} Palladi\={\macron {o}} monit\={\macron {\i}}, pr\={\macron {o}} n\={\macron {u}}mine l{\ae}s\={\macron {o}}}
  { For Pallas he advised, for the wounded pride this that  }
  {110}
  { Having been advised by Calchas, they built this effigy for the Palladium for the offended divine power  }
  { Notes }


\latline
  {effigiem statu\={\macron {e}}re, nef\={\macron {a}}s qu{\ae} tr\={\macron {\i}}ste pi\={\macron {a}}ret.}
  { the build a statue, which sad crime it might atone for. }  
  {110}
  { (of Pallas) which might atone for the sad crime. }
  { Notes }


\latline
  {[\textbf{185}] Hanc tamen imm\={\macron {e}}nsam Calch\={\macron {a}}s attollere m\={\macron {o}}lem}
  { But Calchas commanded that this immense structure, having been woven of oak, be erected }  
  {110}
  { But Calchas commanded them to raise up this immense structure, with oak having been woven } 
  { \begin{enumerate}
  	\item Jussit does not use ``ut'' to make indirect commands.  It uses complimentary infinitives.
  \end{enumerate} }


\latline
  {r\={\macron {o}}boribus text\={\macron {\i}}s c{\ae}l\={\macron {o}}que \={\macron {e}}d\={\macron {u}}cere jussit}
  { and that it be brought up to heaven and that it }     
  {110}
  { and ordered them to bring it up to heaven in order that it }
  { Notes }


\latline
  {n\={\macron {e}} recip\={\macron {\i}} port\={\macron {\i}}s aut d\={\macron {u}}c\={\macron {\i}} in moenia posset,}
  { not be able to be received by the gates or be led into the city walls, }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {neu populum ant\={\macron {\i}}qu\={\macron {a}} sub relligi\={\macron {o}}ne tu\={\macron {e}}r\={\macron {\i}}.}
  { lest it guard the population under the old religion. }
  {110}
  { CorrectedTranslation }
  { Notes }


\latline
  {Nam s\={\macron {\i}} vestra manus viol\={\macron {a}}sset d\={\macron {o}}na Minerv{\ae},}   
  { For if their hands would violate the git of Minerva }          
  {110}
  {He sings that great destruction would come if your hands had violated the gift of Minerva }
  { \begin{enumerate}
  	\item short for violav\={\i}sset
  \end{enumerate} }


\latline
  {[\textbf{190}] tum magnum exitium (quod d\={\macron {\i}} prius \={\macron {o}}men in ipsum}
  { then great destruction (which the gods previously in omen he himself ) }          
  {110}
  { (would that the gods turned this omen   }   
  { \begin{enumerate}
  	\item optative subjunctive:  ``if only / would that / be it that''
  \end{enumerate} }


\latline
  {convertant!) Priam\={\macron {\i}} imperi\={\macron {o}} Phrygibusque fut\={\macron {u}}rum;}
  { had turned), the power of Priam and the Phrygians would be; } 
  {110}
  { against this man ), and for the empire of Priam and the Phrygians; }
  { \begin{enumerate}
  	\item I wish that the gods had punished Calchas only!
  \end{enumerate} }


\latline
  {s\={\macron {\i}}n manibus vestr\={\macron {\i}}s vestram ascendisset in urbem,}      
  { but if they would mount the horse in their city by means of their hands }      
  {110}
  { but if went up into your city (O Trojans), by your hands }
  { Notes }


\latline
  {ultr\={\macron {o}} Asiam magn\={\macron {o}} Pelop\={\macron {e}}a ad moenia bell\={\macron {o}}}
	{ voluntarily they would greatly conquer Asia to the walls of Pelops   }
  {110}
  { voluntarily Asia will come to the walls of Pelops with a great war  }    
  { Notes }


\latline
  {vent\={\macron {u}}ram et nostr\={\macron {o}}s ea f\={\macron {a}}ta man\={\macron {e}}re nep\={\macron {o}}t\={\macron {e}}s.'}
	{ and even these fates would command our grandchildren.'  }
  {110}
  { and these fates would remain for our grandchildren.'  }
  { Notes }


\latline
  {[\textbf{195}] T\={\macron {a}}libus \={\macron {\i}}nsidi\={\macron {\i}}s perj\={\macron {u}}r\={\macron {\i}}que arte Sin\={\macron {o}}nis}
	{ Such evils and the by the perjured art of Sinon }
  {110}
  { By such plots and the art of perjured  Sinon }           
  { Notes }


\latline
  {cr\={\macron {e}}dita r\={\macron {e}}s, capt\={\macron {\i}}que dol\={\macron {\i}}s lacrim\={\macron {\i}}sque co\={\macron {a}}ct\={\macron {\i}}s}
	{ we believed things, and having been captured by treacheries, and having caused tears }
  {110}
  { the matter was believed, and we were taken by treacheries, and by forced tears }
  { Notes }


\latline
  {qu\={\macron {o}}s neque T\={\macron {y}}d\={\macron {\i}}d\={\macron {e}}s noc L\={\macron {a}}r\={\macron {\i}}s{\ae}us Achill\={\macron {e}}s,}
	{ which neither Tydides nor Achilles son of Larisaeus }
  {110}
  { we whom neither Diomedes nor Achilles from Larissa }
  { Notes }


\latline
  {n\={\macron {o}}n ann\={\macron {\i}} domu\={\macron {e}}re decem, n\={\macron {o}}n m\={\macron {\i}}lle car\={\macron {\i}}n{\ae}.}
	{ nor ten years nor a thousand hulls caused. }
  {110}
  { nor ten years, nor a thousand hulls we crushed. }
  { Notes }


\latline
  {H\={\macron {\i}}c aliud majus miser\={\macron {\i}}s mult\={\macron {o}}que tremendum}
	{ Here, to the wretches, another greater omen and an even greater fearful thing }
  {110}
  { Here, to the wretches, another thing, greater and even more fearful thing }
  { Notes }



\latline
  {[\textbf{200}] objicitur magis atque impr\={\macron {o}}vida pectora turbat.}
	{ was presented and was turning our unforeseeing breasts. }
  {110}
  { is presented and was stirring up our unforeseeing breasts. }
  { Notes }


