\latline
  {Ac pr\={\macron {\i}}mum silic\={\macron {\i}} scintillam exc\={\macron {u}}dit Ach\={\macron {a}}t\={\macron {e}}s}
  { And then Achates strikes flint to spark }
  {110}
  { First Achates struck out a spark with a flint }
  { Notes }



\latline
  {susc\={\macron {e}}pitque ignem foli\={\macron {\i}}s atque \={\macron {a}}rida circum}
  { and catches fire on leaves and gives it nearby }
  {110}
  { and caught up fire in the leaves and gives it nearby }
  { Notes }


\latline
  {n\={\macron {u}}tr\={\macron {\i}}menta dedit rapuitque in f\={\macron {o}}mite flamam.}
  { fuel and snatches up the flames in tinder. }
  {110}
  { fuel, and catches it up in the tinder. }
  { Notes }


\latline
  {Tum Cererem corruptam und\={\macron {\i}}s Cere\={\macron {a}}liaque arma}
  { Then they take out grain ruined by vawes, and grain tools, }
  {110}
  { They bring out the wave-corrupted grain and tools of Ceres }
  { Notes }



\latline
  {expediunt fess\={\macron {\i}} r\={\macron {e}}rum, fr\={\macron {u}}g\={\macron {e}}sque recept\={\macron {a}}s}
  { weary of all things and both parch the  }
  {110}
  { weary of all things and the salvaged grain }
  { Notes }


\latline
  {et torr\={\macron {e}}re parant flamm\={\macron {\i}}s et frangere sax\={\macron {o}}.}
  { rescued grain on the flames and grind it on rock. }
  {110}
  { and prepare to roast [it] with flames and break it with rock. }
  { Notes }


\latline
  {Aen\={\macron {e}}\={\macron {a}}s scopulum intere\={\macron {a}} c\={\macron {o}}nscendit, et omnem}
  { Meanwhile Aeneas climbs a rock, searches all views }
  {110}
  { Meanwhile {\AE}neas climbs a cliff and  }
  { Notes }



\latline
  {pr\={\macron {o}}spectum l\={\macron {a}}t\={\macron {e}} pelag\={\macron {o}} petit, Anthea s\={\macron {\i}} quem}
  { far and wide on the see, if he might see anything of wind-thrown }
  {110}
  { seeks a view on the entire see if he may  }
  { \begin{enumerate}
  	\item s\={\i} marks an indirect question
  	\item \textbf{quem} is operating as \textbf{aliquem}, a general supposition per the ``nisi / s\={\i}'' rules
  \end{enumerate} }


\latline
  {jact\={\macron {a}}tum vent\={\macron {o}} videat Phrygi\={\macron {a}}sque bir\={\macron {e}}m\={\macron {\i}}s}
  { Antheus and Phrygian galleys }
  {110}
  { see some wind-tossed Antheus, and if he may see the Phrygian triremes }
  { Notes }


\latline
  {aut Capyn aut cels\={\macron {\i}}s in puppibus arma Ca\={\macron {\i}}c\={\macron {\i}}.}
  { or either Capys or the shields of Caicus on towering decks. }
  {110}
  { CorrectedTranslation }
  { Notes }



\latline
  {N\={\macron {a}}vem in c\={\macron {o}}nspect\={\macron {u}} n\={\macron {u}}llam, tr\={\macron {\i}}s l\={\macron {\i}}tore cerv\={\macron {o}}s}
  { He sees no ships in view, but he sees three stags running on the  }
  {110}
  { No ship in sight, he sees three stags on the beach }
  { Notes }


\latline
  {pr\={\macron {o}}spicit errant\={\macron {\i}}s; h\={\macron {o}}s t\={\macron {o}}ta armenta sequuntur}
  { shore, those, the full herd follows behind, }
  {110}
  { wandering, those stags, the whole herd follows from behind,  }
  { Notes }


\latline
  {\={\macron {a}} terg\={\macron {o}} et longum per vall\={\macron {\i}}s pascitur agmen.}
  { distantly, across the valley and they graze as a troop }
  {110}
  { distantly, across the valley, and grazes }
  { Notes }



\latline
  {C\={\macron {o}}nstitit h\={\macron {\i}}c arcumque man\={\macron {u}} celer\={\macron {\i}}sque saggit\={\macron {a}}s}
  { Here Aeneas sets and snatches both bow and arrows with }
  {110}
  { He sets place here and grabbed with his hand the bow and swift arrows  }
  { \begin{enumerate}
  	\item C\={o}nstitit is \textbf{perfect}
  \end{enumerate} }


\latline
  {corripuit, f\={\macron {\i}}dus quae t\={\macron {e}}la ger\={\macron {e}}bat Ach\={\macron {a}}t\={\macron {e}}s,}
  { a swift hand, weapons which faithful Achates had brought  }
  {110}
  { which faithful Achates was carrying }
  { Notes }


\latline
  {duct\={\macron {o}}r\={\macron {e}}sque ips\={\macron {o}}s pr\={\macron {\i}}mum capita alta ferent\={\macron {\i}}s}
  { and he lays low those very leaders with branching horns, then he lays low the herd  }
  {110}
  { and he lays low first those leaders themselves, carrying their high heads with branching  }
  { Notes }



\latline
  {cornibus arbore\={\macron {\i}}s sternit, tum vulgus et omnem}
  { he then attacks the herd and  }
  {110}
  { horns and then mixes the entire herd }
  { Notes }


\latline
  {miscet ag\={\macron {e}}ns t\={\macron {e}}l\={\macron {\i}}s nemora inter frondea turbam;}
  { confuses all, sending the confusion into the leafy grove with his weapons }
  {110}
  { flushing them into the leafy grave with his weapons }
  { Notes }


\latline
  {nec prius absistit quam septem ingentia victor }
  { not stopping until the victor gathers seven huge }
  {110}
  { And he didn't stop before the victor slays }
  { \begin{enumerate}
  	\item prius{\ldots}quam is ``\textbf{priusquam}'' which is always separated
  \end{enumerate} }



\latline
  {corpora fundat hum\={\macron {\i}} et numerum cum n\={\macron {a}}vibus aequet.}
  { carcasses on the ground, he equals the count with the number of his ships. }
  {110}
  { 7 huge carcasses to the earth and he makes equal the number to the ships. }
  { Notes }


\latline
  {Hinc portum petit et soci\={\macron {o}}s part\={\macron {\i}}tur in omn\={\macron {\i}}s.}
  { He finds the port and divides them among all his compatriots. }
  {110}
  { Then he seeks the port and divides them among all his compatriots. }
  { Notes }


\latline
  {V\={\macron {\i}}na bonus quae deinde cad\={\macron {\i}}s oner\={\macron {a}}rat Acest\={\macron {e}}s}
  { Good Acestes unloads Sicilian wine jars on the beach }
  {110}
  { He divides wine which good Acestes had  }
  { \begin{enumerate}
  	\item oner\={a}rat is poetic ``oneraverat'', thus perfect
  \end{enumerate} }



\latline
  {l\={\macron {\i}}tore Tr\={\macron {\i}}nacri\={\macron {o}} dederatque abeuntibus h\={\macron {e}}r\={\macron {o}}s}
  { and gives to departed heroes and [then] divided, and calmed their grieving hearts with words. }
  {110}
  { given on the Sicilian shore to the departing heroes }
  { Notes }