All were silent and held their mouths intently.
Afterward, from his high couch, father {\AE}neas speaks:
``It is a wicked thing, queen, the sadness you order to relive,
how the Dana\"ans overthrew Trojan wealth and the sad kingdom
although I myself saw these most miserable things
and I was a large part of them.  Who having heard these thins
of the Myrmidonians or the Dolupians or what soldier of
hard Ulysses might hold himself from tears?  And now already
the humid night precipitates and the falling stars
urge sleep.
But if your desire to know our sad fates and
to briefly hear the final labor of Troy is so great,
although my soul hakes remembering it and
flees from the grief, I begin:

The men, broken by war, and the repulsed Dana\"an leaders,
with the years rolling by, build
a horse structure, big as a mountain, with the divine
art of Minerva, and weave, fir having been cut, ribs;
they feign it a prayer for the return; their reputation
is spread by this thing.
Here, the bodies of men chosen by lot are enclosed
in dark flanks and within
gigantic caverns they fill each with equipped soldiers.

It is with Tenedos in sight, an island most known in reputation,
rich in resources while the kingdom of Priam remained,
now but a channel and anchorage of ill faith to hulls:
hither they are taken away and conceal themselves on a
deserted beach.  We thought that they
had departed and sought Mycenae with the wind.
Therefore all free themselves from the grief of the Teucrians:
the gates are opened, and it pleases us to go into
the deserted Doric camps and to see the places and
the abandoned beach:
here were the Dolopian bands, here stretched savage Achilles;
here is the place for ships; here the battle-lines were used to fighting.
A portion of the virgins are stupefied by the fatal gift of Minerva and
gaze at the teeth of the horse; Thymoetes firsts urges
that the horse be taken inside the walls and set within the citadel;
or if whether by some trick the fates seek to carry off Troy.
But Capys, and those of better plans in the mind order that
the Dana\"an plots and suspicious gifts be thrown headlong to the sea and
be burned up with flames thrown thereunder or also that holes
be bored into the uterus and the hiding places be probed.
Unclear zeal splits the crowd in converse.

First, there, before all, a great crowd accompanying, burning Laocoon runs in the great noisy mob from the
height of the citadel and says out far: `O miserable citizens, what such great insanities
are these?   Do you trust that the enemies have then departed?  Do you reckon any gifts from the
Dana\"ans to be free of treacheries.  Is Ulysses known thus?  Either in this wood concealed
Argives are hidden, or this machine has been made against our walls to inspect our homes
or to come from above to the city, or some
other error lies here; do not trust the horse, Teucrians.
Whatever it is, I fear Greeks, especially those bringing gifts.'
Thus he spoke with true intentions and threw
a gigantic spear in the flank, into the curved joints of the belly of the beast.  That spear 
stood there, shaking, and as the uterus was hit, the caves sounded and the caverns gave a groan .
And, if the fates of gods, if the mind had not been hostile, he would have push to
befoul the Argive nests with iron, and troy would now stand, and
the high citadel of Priam would now remain.

Behold, Dardan pastors are dragging a youth with his hands bound behind his 
back with a great clamor to the king, a youth who voluntarily carried
himself as a stranger to the arriving men and who by this very act would build Troy and open it to the 
Argives.  A youth, true in his soul, and ready for either case: whether to turn over with tricks or to
fall to certain death.

All around, poured-out Trojan youth curious to see rushes and
mock the captive.  Take in now the treacheries
of the Dana\"ans and by the crime of one, know all.

And now as seen in the middle of the crowd, jostled, unarmed,
he stood and locked around the force of Phrygia with his eyes:
`Alas, what land to me,' he began `what sea is possible to accept me?
or what misery remains at last to me, me to whom is neither
a place among the Dana\"ans and besides from whom the hostile
Dardans themselves demand repayment of crimes by blood?'

By means of his cry the wraths are stayed
and all violence checked.  We urge him to tell by what blood he
was born, or what news he carries; let him recount what things he 
hoped for as a captive.
[That man relates these things even as his fear is dropped:]